# Reanalysis 


```{r}
#| echo: false
#| include: false
#| purl:  false
source("scripts/before_chapter_script.R")
```

A reanalysis of an existing study is a follow-up study that reuses the original realized data for some new purpose. The reanalysis is a study with a research design that can be described in terms of *M*, *I*, *D*, and *A*. Reanalyses are fundamentally constrained by the data strategy of the original study. The data strategy *D* and the resulting data are set in stone -- but reanalysts can make changes to the answer strategy *A* and sometimes also to the model *M* or inquiry *I*.

We can learn from reanalyses in several ways. First, we can fix errors in the original answer strategy. Reanalyses fix simple mathematical errors, typos in data transcription, or failures to account for features of the data strategy when analyzing the data. These reanalyses show whether the original results do or do not depend on these corrections. Second, we can reassess the study in light of new information about the world learned after the original study was published. That is, sometimes *M* changes in ways that color our interpretation of past results. Perhaps we learned about new confounders or alternative causal channels that undermine the original design's credibility. When reanalyzed, demonstrating the results do (or do not) change when new model features are incorporated improves our understanding of the inquiry. Third, reanalyses may also aim to answer new questions that were not considered by the original study but for which the realized data can provide useful answers. 

Lastly, many reanalyses show that original findings are not "robust" to alternative answer strategies. These are better conceptualized as claims about robustness to alternative models: one model may imply one answer strategy, and a different model, with another confounder, suggests another. If both models are plausible, a good answer strategy should be robust to both and even help distinguish between them. A reanalysis could uncover robustness to these alternative models or lack thereof. 

Reanalyses are themselves research designs. Just like any design, whether a reanalysis is a strong research design depends on *possible* realizations of the data (as determined by the data strategy), not just the realized data. Because the realized data is fixed in a reanalysis, analysts are often instead tempted to judge the reanalysis based on whether it overturns or confirms the original study's results. A successful reanalysis in this way of thinking demonstrates, by showing that the original results are changed under an alternative answer strategy, that the results are not robust to other plausible models. 

This way of thinking can lead to incorrect assessments of reanalyses. Instead, we should consider what answers we would obtain under the original answer strategy *A* and the reanalysis strategy *A'* under many *possible* realizations of the data. A good reanalysis strategy reveals with high probability the set of models of the world under which we can make credible claims about the inquiry. Whether or not the results change under the answer strategies *A* and *A'* tells us little about this probability because the realized data is only one draw. 

### Example

In this section, we illustrate the flaw in assessing reanalyses based on changing significance of results alone. We demonstrate how to assess the properties of reanalysis plans, comparing the properties of original answer strategies to proposed reanalysis answer strategies. 

The design we consider is an observational study with a binary treatment $Z$ that may or may not be confounded by a covariate $X$. Suppose that the original researcher had in mind a model in which $Z$ is not confounded by $X$:

```{r, file = "scripts_declarations/declaration_23.1a.R"}
```

The reanalyst has in mind a different model. In this second model, $X$ confounds the relationship between $Z$ and $Y$:

```{r, file = "scripts_declarations/declaration_23.1b.R"}
```

The original answer strategy `A` is a regression of the outcome $Y$ on the treatment $Z$. The reanalyst collects the covariate $X$ and proposes to control for it in a linear regression; call that strategy `A_prime`. 

```{r, file = "scripts_declarations/declaration_23.1c.R"}
```

