# Cluster-randomized experiments {.unnumbered}

```{r}
#| echo: false
#| include: false
#| purl:  false
source("scripts/before_chapter_script.R")
```

:::: {.ddbox}
We declare a cluster-randomized trial in which subjects are assigned to treatment and control conditions in groups. We use design diagnosis to quantify how the magnitude of the efficiency losses from clustering depend on the intra-cluster correlation.
::::

When whole groups of units are assigned to treatment conditions together, we say that the assignment procedure is clustered. A common example is an education experiment in which the treatment randomized at the classroom level. All students in a classroom are assigned to either treatment or control together; assignments do not vary within classroom. Clusters can be localities, like villages, precincts, or neighborhood. Clusters can be households if treatments are assigned at the household level. 

Typically, cluster randomized trials exhibit higher variance than the equivalent individually-randomized trial. How much higher variance depends on a statistic that can be hard to think about, the intra-cluster correlation (ICC) of the outcome. The total variance can be decomposed into the variance of the cluster means $\sigma^2_{\textrm{between}}$ plus the individual variance of the cluster-demeaned outcome $\sigma^2_{\textrm{within}}$. The ICC is a number between zero and one that describes the fraction of the total variance that is due to the between variance: $ICC = \frac{\sigma^2_{\textrm{between}}}{\sigma^2_{\textrm{between}} + \sigma^2_{\textrm{within}}}$. If ICC equals one, then all units within a cluster express the same outcome, and all of the variation in outcomes is due to cluster-level differences. If ICC equals zero, then the cluster means are all identical, but the individuals vary within each cluster. When ICC is one, the effective sample size is equal to the number of clusters. When ICC is zero, the effective sample size is equal to the number of individuals. Since ICC is usually somewhere between these two values, we can see that clustering decreases the effective sample size from the number of individuals. The size of this decrease depends on how similar outcomes are within cluster compared to how similar outcomes are across clusters.

For these reasons clustered random assignment is not usually a desirable feature of a design. Sometimes, however, it is useful or even  *necessary* for logistical or ethical reasons for subjects to be assigned together in groups. 

To demonstrate the consequences of clustering, Declaration @def-declarationch18num5 shows a design in which both the untreated outcome `Y_Z_0` and the treatment effect `tau_i` exhibit intra-cluster correlation. The inquiry is the average treatment effect over individuals which can be defined without reference to the clustered structure of the data. The data strategy employs clustered random assignment. We highlight two features of the clustered assignment. First, the clustered nature of the data does not itself *require* for clustered assignment. In principle, one could assign treatments at the individual level or subgroup level even if outcomes are correlated within groups. Second, surprisingly, random assignment of clusters to conditions does not guarantee unbiasedness of outcomes when clusters are of unequal size [@middleton2008bias;  @imai2009essential]. The bias stems from the possibility that potential outcomes could be correlated with cluster size. With uneven cluster sizes, the total number of units (the denominator in the mean estimation) in each group bounces around from assignment to assignment. Since the expectation of a ratio is not, in general, equal to the ratio of expectations, any dependence between cluster size and potential outcomes will cause bias. We can address this problem by blocking clusters into groups according to cluster size. If all clusters in a block are of the same size, then then overall size of the treatment group will remain stable from assignment to assignment. For this reason the design below uses clustered assignment blocked on cluster size.

:::{.definition #declarationch18num5} 

Blocked and clustered randomized trial

```{r, file = "scripts_declarations/declaration_18.5.R"}
```

:::


:::{.lemma #diagnosisch18num5}

Redesigning over values of ICC

```{r}
#| eval: false

diagnosis_18.5 <-
  declaration_18.5 |>
  redesign(ICC = seq(0.1, 0.9, by = 0.4)) |> 
  diagnose_designs()
```

Figure @fig-ch18num5 shows the sampling distribution of the difference-in-means estimator under cluster random assignment at three levels of intra-cluster correlation ranging from 0.1 to 0.9.

The top row of panels plots the treatment effect on the vertical axis and the untreated potential outcome on the horizontal axis. Clusters of units are circled. At low levels of ICC, the circles all overlap, because the differences across clusters are smaller than the differences within cluster. At high levels of ICC, the differences across clusters are more pronounced than differences within cluster. The bottom row of panels shows that the sampling distribution of the difference-in-means estimator spreads out as the ICC increases. At low levels of ICC, the standard error is small; at high levels the standard error is high.

![Sampling distribution under differenct ICCs](/figures/figure-18-5){#fig-ch18num5}

This diagnosis clarifies the *costs* of cluster assignment. These costs are greatest when there are few clusters and when units within clusters have similar potential outcomes. Diagnosis can be further used to compare these costs to advantages and assess the merits of variations in the design that seek to alter the number or size of clusters.

:::

## Design examples

- @mousa2020building studies the effects of inter-group contact on tolerance by cluster assigning players in a Christian football league in Iraq to play with four new Muslim teammates or four new Christian teammates, where the clusters are the teams. 

- @paluck_green_2009 cluster assigned communities in Rwanda to radio programs encouraging dissent and disobedience to authorities and measured individual-level outcomes via survey.

