# Getting started {#sec-ch4}

```{r}
#| echo: false
#| include: false
#| purl:  false
source("scripts/before_chapter_script.R")
```


This chapter serves as a quick start guide for the code used throughout this book, and in particular the `DeclareDesign` package for the R programming language. `DeclareDesign` is a software implementation of every step of the declare-diagnose-redesign process. While you can declare, diagnose, and redesign using nearly any programming language, `DeclareDesign` is structured to make it easy to mix and match design elements while handling the tedious simulation bookkeeping behind the scenes. 

First, we provide instructions for getting started in R and RStudio. We then introduce the code structure for the three steps of the research planning process: declaration, diagnosis and redesign. After this introduction, readers should be able to *use* the code, but perhaps not write it themselves yet. We devote a longer section (Chapter @sec-ch13) to getting started *writing* the code (@sec-ch13).

## Installing R 

You can download R for free from [CRAN](https://cran.r-project.org). We also recommend the free program [RStudio](https://rstudio.com/products/rstudio/download), which provides a friendly interface to R. Both R and RStudio are available on Windows, Mac, and Linux.

Once you have R and RStudio installed, open up RStudio and install `DeclareDesign` and its related packages. These include three packages that enable specific steps in the research process: `fabricatr` for simulating social science data, `randomizr` for random sampling and random assignment, and `estimatr` for design-based estimators. You can also install `rdss`, which includes datasets and helper functions used in the book. To install them all, copy the following code into your R console:

```{r install_package_example, eval = FALSE}
install.packages(c("DeclareDesign", "rdss"))
```

We also recommend that you install and get to know the `tidyverse` set of packages for data analysis, which we will use throughout:

```{r install_tidyverse_example, eval = FALSE}
install.packages("tidyverse")
```

For introductions to R and the `tidyverse` we especially recommend the free resource [*R for Data Science*](https://r4ds.had.co.nz).

All of the code in this book assumes that the `DeclareDesign` family of packages, the `tidyverse` suite, and the companion package to the book `rdss` have been loaded. Once they have been installed on your computer, you can load them with the following code:

```{r}
library(DeclareDesign)
library(tidyverse)
library(rdss)
```

## Declaration

Designs are constructed from design elements: models, inquiries, data strategies, and answer strategies. 

In `DeclareDesign`, each design element is made with a function that starts with the word `declare`. For example, we can declare an assignment procedure using `declare_assignment` as follows:

```{r}
simple_random_assignment <- 
  declare_assignment(Z = simple_ra(N = N, prob = 0.6))
```

Each element created by a `declare_*` function, perhaps surprisingly, is itself a function. The object `simple_random_assignment` is not a particular assignment --- instead, it is a function that conducts assignment when called. Each time we call `simple_random_assignment` we get a different random assignment:

```{r}
#| eval: false

participants <- data.frame(ID = 1:100)

assignment_1 <- simple_random_assignment(participants)
assignment_2 <- simple_random_assignment(participants)
assignment_3 <- simple_random_assignment(participants)

bind_cols(assignment_1, assignment_2, assignment_3)
```
```{r}
#| echo: false
#| label: tbl-multiple-assignments
#| tbl-cap: Three random assignments from the same random assignment step.
#| tbl-subcap: ["1", "2", "3"]
#| layout-ncol: 3

participants <- data.frame(ID = 1:100)

set.seed(7)

assignment_1 <- simple_random_assignment(participants)
assignment_2 <- simple_random_assignment(participants)
assignment_3 <- simple_random_assignment(participants)
head(assignment_1, 5) |> 
  kable(booktabs = TRUE, align = "c", digits = 3) |> 
  kable_styling()
head(assignment_2, 5) |> 
  kable(booktabs = TRUE, align = "c", digits = 3) |> 
  kable_styling()
head(assignment_3, 5) |> 
  kable(booktabs = TRUE, align = "c", digits = 3) |> 
  kable_styling()
```

Every step in a research design can be declared using one of the `declare_*` functions. @tbl-declarationfunctions collects these according to the four elements of a research design. In @sec-ch13, we detail how to build each kind of step.

| Design component | Function                       | Description                                   |
| ---------------- | ------------------------------ | --------------------------------------------- |
| Model            | `declare_model()`              | background variables and potential outcomes   |
| Inquiry          | `declare_inquiry()`            | research questions                            |
| Data strategy    | `declare_sampling()`           | sampling procedures                           |
|                  | `declare_assignment()`         | assignment procedures                         |
|                  | `declare_measurement()`        | measurement procedures                        |
| Answer strategy  | `declare_estimator()`          | estimation procedures                         |
|                  | `declare_test()`               | testing procedures                            |

: Declaration functions in DeclareDesign {#tbl-declarationfunctions} 

We use the `+` operator to build from elements of a design to a design. @def-ch4num1 shows the format of most declarations throughout the book. This declaration represents a two-arm randomized experiment with 100 units from which we aim to estimate the average treatment effect.

::: {#def-ch4num1 .declaration} 


$~$

:::


## Diagnosis {#diagnose-design-function}

Diagnosis is the process of simulating the design many times and calculating summary statistics about the design that describe its properties, which we call diagnosands. Once a design is declared, diagnosis is as simple as using the `diagnose_design` function on it.

::: {#lem-ch4num1} 

Example design diagnosis
Two-arm randomized experiment

![Two-arm randomized experiment declaration](/figures/declaration-example){#fig-designannotated}

```{r}
#| eval: false
diagnose_design(declaration_4.1, sims = 100)
```



:::

The output of the diagnosis includes the diagnosand values (top row), such as bias of $-0.02$, and our uncertainty about the diagnosand value (bootstrapped standard error in parentheses in the bottom row). The uncertainty estimates tell us whether we have conducted enough simulations to precisely estimate the diagnosands. The fact that that the estimate of bias is $-0.02$ and the standard error is $0.03$ means that we cannot distinguish the amount of bias from no bias at all. 

```{r}
#| tbl-cap: "Design diagnosis."
#| echo: false
#| label:  tbl-diagnosis4s1
diagnosis_4.1 <- get_rdddr_file_local("diagnosis_4.1")

diagnosis_4.1 |>
  reshape_diagnosis() |>
  select(Bias, RMSE, Power) |>
  kable(
    booktabs = TRUE,
    align = "c",
    digits = 3
  ) |> 
  kable_styling()
```

## Redesign {#redesign-function}

We redesign to learn how the diagnosands change as design features change. We can do this using the `redesign` function over a range of sample sizes, which produces a list of designs.

```{r}
#| eval: false

designs <- redesign(declaration_4.1, N = c(100, 200, 300, 400, 500))
```

Our simulation and diagnosis tools can operate directly on this list of designs:

```{r}
#| eval: false

diagnose_design(designs)
```

## Library of designs

In our `DesignLibrary` package, we have created a set of common designs as "designers" (functions that create designs from just a few arguments), so you can get started quickly.

```{r}
#| eval: false

library(DesignLibrary)

block_cluster_design <- 
  block_cluster_two_arm_designer(N = 1000, N_blocks = 10)
```

## Long term code usability 

We have written the code examples with`DeclareDesign` version 1.0.0, the package version we released along with the book. We are committed to the long-term maintenance of this software, but inevitably, the evolution of the `R` ecosystem and further package development will mean that some of the printed code will break in the future. 

However, even if code eventually becomes obsolete, a virtue of writing out designs in code is that they are explicit: the entire design is encoded in the declarations we provide. Even if the code itself won't run, you can still use it to understand the design and to draw insights from the diagnosis.

<!-- ## Further reading {-} -->

<!-- This primer is an introduction to DeclareDesign, but it only scratches the surface of what we can do with the software. At the end of each section in Part II, we illustrate how to tackle interesting problems that come up during declaration, diagnosis, and redesign. -->

<!-- In the meantime, we recommend the following resources for learning more. -->

<!-- - [DeclareDesign.org](declaredesign.org) -->
<!-- - [randomizr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/randomizr.pdf) -->
<!-- - [estimatr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/estimatr.pdf) -->
<!-- - [DeclareDesign cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/declaredesign.pdf) -->
<!-- - [fabricatr vignette](https://declaredesign.org/r/fabricatr/)  -->
<!-- - [R for Data Science](https://r4ds.had.co.nz) -->
<!-- - [RStudio R primers](https://rstudio.cloud/learn/primers) -->
<!-- - [Computational social science bootcamp](https://sicss.io/boot_camp) -->
